---
title: "as5457_final"
author: "Sunny Siddique"
date: "November 29, 2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
```

```{r}
HIV_data = read_csv("./data/DOHMH_HIV_AIDS_Annual_report.csv")
```

cleaning the dataset for additonal analysis...

```{r}
hiv = HIV_data %>% 
  janitor::clean_names(dat = .)%>%
  filter(uhf != "All")%>%
  filter(borough != "All")%>%
  filter(gender != "All")%>%
  filter(age != "All")
```

Are there differences in HIV diagnosis rate between boroughs?
```{r}
fit = lm(hiv_diagnosis_rate ~ borough, data=hiv)
summary(fit)
summary(fit)$coef
```

Trying to tidy up the output
```{r}
fit %>% 
  broom::tidy() %>% 
  mutate(term = str_replace(term, "^borough", "Borough: ")) %>%
  knitr::kable(digits = 3)
```

Compared to the Bronx, all neighborhoods have statistically significant infection rates. 

Is age a significant predictor of hiv diagnosis rate?
```{r}
fit1 = lm(hiv_diagnosis_rate ~ age, data = hiv)
fit1 %>% 
  broom::tidy() %>% 
  knitr::kable(digits = 3)
```

Compared to 13-19 age group, 20-29 year olds have 42.6 more HIV diagnoses on average. 

What about race?
```{r}
race_clean = HIV_data %>% 
  janitor::clean_names() %>% 
  filter(race != "All") %>% 
  mutate(race = forcats::fct_relevel(race, c("White", "Black", "Latino/Hispanic", "Asian/Pacific Islander", "Other/Unknown")))

fit2 = lm(hiv_diagnosis_rate ~ race, data = race_clean)
fit2 %>% 
  broom::tidy() %>% 
  knitr::kable(digits = 3)
```

Compared to Whites, all other racial groups have a significantly different HIV diagnosis rate. 

let's fit in a simple linear regression model
```{r}
lin_reg = HIV_data %>%
  janitor::clean_names() %>% 
  filter(borough != "All")%>%
  filter(gender != "All")%>%
  filter(age != "All")
```

```{r}
fit3 = lm(hiv_diagnosis_rate ~ borough + gender + age, data = hiv)
linear_output = fit3 %>% 
  broom::tidy() %>% 
  knitr::kable(digits = 3)
linear_output
```

