---
title: "Big Picture"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
---

###Introduction

In the continued fight against HIV/AIDS, New York City is instrumental in upholding the New York state plan to ends the AIDS epidemic by 2020. To guide this progress, we must be able to understand both the level of burden and the improvements made over time.

This page is dedicated to visualizations that highlight big picture analyses  of HIV's impact on New York City:

```{r setup, include=FALSE}
#Loading libraries:

library(tidyverse)
library(plotly)

theme_set(theme_bw() + theme(legend.position = "bottom"))

#Loading the data:
  
hiv_data = read_csv("./data/DOHMH_HIV_AIDS_Annual_Report.csv") %>% 
  janitor::clean_names()
```

### Death Rates (In Relation to HIV) Over the Years

How much HIV-related causes contribute to the overall death rate for people living with HIV in New York City over the years can illustrate how the combined effects of implemented interventions, policies, and strategies have altered the impact of HIV.

The following stacked bar graph highlights the extent to which deaths related to HIV/AIDS contribute to the overall number of deaths per 1,000 people, mid-year, living with HIV/AIDS. Changes in the contribution of HIV-related deaths are shown from the years 2011 to 2014.

```{r death rates by type, echo = FALSE, message=FALSE, warning=FALSE}
#Original code by Bryce, further edited by Bryce

#Cleaning the data to just look at times, controlling for gender, age, race, borough, and UHF:

hiv_year_data = hiv_data %>% 
  filter(
    borough == "All",
    uhf == "All",
    gender == "All",
    age == "All",
    race == "All"
  )

#Making a stacked bar graph in plot.ly

stacked_death_plot = hiv_year_data %>% 
  filter(non_hiv_related_death_rate != 99999,
         hiv_related_death_rate != 99999) 

stacked_death_plot %>% plot_ly(x = ~year, y = ~non_hiv_related_death_rate, type = 'bar', name = 'Non HIV-related death rate') %>%
  add_trace(y = ~hiv_related_death_rate, name = 'HIV-related death rate') %>%
  layout(title = "Death Rate Types over the Years", 
         yaxis = list(title = 'Rate'), 
         barmode = 'stack')
```

There is a downward trend in the total number of deaths per 1,000 people living with HIV/AIDS in New York City over the years. When further exploring the breakdown of causes of death within this overall death rate, it is revealed that the driving component for this reduction in total death rate is the decrease in number of deaths attributed to HIV/AIDS. By contrast, the number of deaths attributed to other causes has remained relatively consistent across the years. This reflects achievements in improving the extent that HIV/AIDS has been fatal for New York City residents living with HIV/AIDS. 

###HIV Diagnosis Rates by Borough

In addressing how to further direct HIV/AIDS efforts in New York City, it is particularly useful to be able to identify how the impact of HIV is distributed across the city's geographic regions. Taking a look at how this distribution has changed over time may also serve to highlight disparities, and thus where additional attention and resources need to be allocated.   

In the bar graph below, we assess the number of HIV diagnoses per 100,000 NYC residents within each borough, between the years 2011 and 2015.

```{r HIV diagnosis rates by boro, echo = FALSE, message=FALSE, warning=FALSE}
#Original code by Noah, edited by Bryce

#Filtering original data to look at summary observations across demographic variables for each borough

HIV_diagnosis_data = hiv_data %>%
  filter(uhf %in% c("All")) %>%
  filter(borough != "All") %>%
  filter(gender == "All") %>%
  filter(age == "All")

#Plotting a bar graph (in plot.ly) looking at diagnoses rate each year for each borough

spread_boros = HIV_diagnosis_data %>% 
  spread(key = borough, value = hiv_diagnosis_rate) %>% 
  janitor::clean_names()

spread_boros %>% plot_ly(x = ~year, y = ~manhattan, type = 'bar', name = 'Manhattan') %>%
  add_trace(y = ~bronx, name = 'Bronx') %>%
  add_trace(y = ~brooklyn, name = 'Brooklyn') %>%
  add_trace(y = ~queens, name = 'Queens') %>%
  add_trace(y = ~staten_island, name = 'Staten Island') %>%
  layout(title = "HIV Diagnosis Rate by Borough over Time",
         yaxis = list(title = 'HIV Diagnosis Rate'), 
         barmode = 'group')
```

HIV diagnoses are not equally distributed between boroughs. Across the years, Manhattan has consistently seen the highest rate of HIV diagnoses per 100,000 residents, followed by the Bronx, then Brooklyn, then Queens, and finally Staten Island. Of note, however, are the downwards trends in diagnosis rate for all three of the most disproportionately-impacted boroughs. Meanwhile, rates in diagnosis have remained relatively stable within Queens and Staten Island. 

While the relative ranking in diagnosis rate has also remained generally consistent from 2011-2015 (e.g., Manhattan is usually the borough with the highest level of HIV diagnoses), 2015 saw Manhattan diagnosing fewer cases of HIV than the Bronx for the first time. This, along with dropping rates for Bronx and Brooklyn, imply improvements amongst these three heavily-burdened areas. Unfortunately, the lack of improvement amongst the remaining boroughs might reveal gaps in progress for certain geographic areas.

###Conclusion

The overall burden caused by HIV/AIDS in New York City has been reduced over the years. This is seen in the improving health outcomes of HIV/AIDS city-wide, as well as in the declining numbers of new infections amongst the most impacted geographic areas. 

Unfortunately, there appears to be evidence of fewer increases in progress amongst less impacted regions. This demonstrates the need to reassess the aims and targets of continuing intervention efforts. 