---
title: "BJL2150_Exploratory_Data_Analysis"
author: "Briana Lettsome"
date: "November 19, 2018"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(ggridges)
```

```{r}
# Reading in of the DOHMH_HIV_AIDS_Annual_Report csv file.

hiv_data = read_csv(file = "./data/DOHMH_HIV_AIDS_Annual_Report.csv") %>%
  janitor::clean_names()
```


```{r}
hiv = hiv_data %>%
  filter(borough == "All") %>%
  select(year, race, hiv_diagnosis_rate, gender) %>%
  filter(!race == "All")

 # Spaghetti Plot showing the HIV Diagnosis Rates per race, faceted by gender.

hiv %>%
  ggplot(aes(x = year, y = hiv_diagnosis_rate, color = race, group = race)) +
  geom_line() +
  facet_grid(~gender) +
   labs(
    title = "HIV Diagnosis Rates by Race",
    x = "Year",
    y = "HIV Diagnosis Rates",
    caption = "Data from the NYCDOHMH"
  )
```

Other/Unknown category includes people of Native American, multiracial, and unknown races.


# PLWDHI prevalence
```{r}
  race_prevalence = hiv_data %>%
    select(age,plwdhi_prevalence, year, uhf, gender, borough, race) %>%
    filter(uhf =="All") %>%
  filter(!race == "All") %>%
  filter(borough == "All") %>%
  select(-age, -uhf, -borough)

 two_race_prevalence = hiv_data %>%
    select(age,plwdhi_prevalence, year, uhf, gender, borough, race) %>%
    filter(uhf =="All") %>%
  filter(!race == "All") %>%
  filter(borough == "All") %>%
  select(-age, -uhf, -borough)

  
  ggplot(race_prevalence, aes(x = year, fill = race)) + 
  geom_histogram(position = "dodge", binwidth = 2) +
    facet_grid(~ gender)
  
race_prevalence %>%
  ggplot(aes(x = year, y = plwdhi_prevalence, color = race)) + 
  geom_line() +
  facet_grid(~gender)

```

## Prevalence by borough

Making of spaghetti plot showing persons diagnosed HIV infection by borough over
the years.
```{r}
 borough_prevalence = hiv_data %>%
    select(age,plwdhi_prevalence, year, borough, uhf, gender, race) %>%
    filter(uhf =="All") %>%
   filter(race == "All") %>%
   filter(gender == "All") %>%
   filter(!borough == "All") %>%
  group_by(year, borough)

borough_prevalence %>%
  ggplot(aes(x = year, y = plwdhi_prevalence, color = borough)) + 
  geom_line() 
```

EDAs to keep:

Bryce - 
Bar graph - Death rate type over the years
Death rate line graph - Will talk about in paper but not include
Percent viral suppression by race and faceted by gender

Briana-
HIV diagnosis rates by race and faceted by gender (can link this with Bryce’s percent viral suppression by race and gender)

Noah -
HIV diagnosis by borough

David -
HIV diagnosis rates by age and faceted by gender


