---
title: "Data Science I: Final Report"
date: "December 6th, 2018"
output: github_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(ggridges)

hiv_data = read_csv(file = "./data/DOHMH_HIV_AIDS_Annual_Report.csv") %>%
  janitor::clean_names()
```

**Project Title**

Identification And Analysis of Trends And Disparities In HIV/AIDS For NYC

**Group Members**

* Bryce Wong (baw2160)
* Sunny Siddique (as5457)
* David Yu (ky2288)
* Noah Kreski (Ntk2109)
* Briana Lettsome (Bjl2150)


**Motivation / Related Work**

According to the the past decade or so, improvements in drug therapies and prevention intervations have led to a decreased number of AIDS cases. In June, 2014, New York Governor Andrew Cuomo announced a new strategy designed to "End the Epidemic" in the state through a combination of increased HIV screening and testing, promotion of Pre-exposure Prophylaxis (PrEP) and improved linkage and retention in clinical care for HIV-positive persons. Since then, the New York City Department of Health and Mental Hygiene (DOHMH) reported a decline in new infections of nearly 8 percent annually or approximately 15% since the campaign was launched.

Despite efforts to reduce the number of new infections, the HIV Surveillance Annual Report presented a number of key points that motivated our decision to analyze HIV infections data. Amongst them, the most noteworthy are:

* Poor New Yorkers, as well as African-American and Hispanic New Yorkers, bear a disproportionate burden of illness and premature death.
* Poor health is concentrated in certain New York City neighborhoods.
* Factors associated with poor health, such as poor access to medical care, unhealthy behaviors, and poor living conditions, are more common among certain economic and racial/ethnic groups.

**Initial Questions**

* Are there disparities in HIV infections / diagnoses based on NYC neighborhoods, if so, which neighborhoods have the highest incidence of HIV infections? Which neighborhoods have the lowest incidence?
* Are there disparities in common demographic factors, including age, gender and race? 
* Does HIV related death rate appear elevated for certain social groups? What about non-HIV related deaths?
* Are there specific observable trends in "percent viral suppression" over the past few years? If so, do the trends appear comparable to the DOHMH interventions following Governor Cuomo's announcements in 2014?



**Methods**

Data: Source, scraping method, cleaning, etc.

  The data that was utilized for this project was the New York City Department of
Health and Mental Hygeine HIV/AIDS Surveillance Annual Report. The dataset was 
downloaded from NYCOpen Data, an open website that provides information and data to 
New Yorkers by the NYC government, an imported into R as a csv file. 
  Within this particular datThe data contained within the dataset were collected between the years 2011 through 2016. The HIV/AIDS surveillance Annual Report contains eighteen variables, including
year, age, race, HIV diagnoses, deaths, and percent viral suppression. The variables used within the visualizations are a followed:





**Exploratory analysis**

: Visualizations, summaries, and exploratory statistical analyses. Justify the steps you took, and show any major changes to your ideas.

**Big Picture Visualizations**

```{r}
# Developing the code to visually assess HIV diagnosis rates over time by borough

HIV_diagnosis_data = hiv_data%>%
               janitor::clean_names()%>%
               filter(uhf %in% c("All"))%>%
               filter(borough != "All")%>%
               filter(gender == "All")%>%
               filter(age == "All")

# Creating a plot to visually assess the above dataframe.

HIV_diagnosis_data%>%
            ggplot(aes(x = year, y = hiv_diagnosis_rate, group = borough, 
                       color = borough))+geom_point() + geom_line() +
  labs(
    title = "HIV Diagnosis Rate by Borough Over time",
    x = "Year",
    y = "HIV Diagnosis Rates",
    caption = "Figure 1. Data from the NYCDOHMH HIV/AIDS Surveillance Annual Report"
  )
```

   In Figure 1, The original dataset was filtered down to get the summary data by
borough for each year. Then, a bar plot was used for the HIV diagnosis rates,
clustered by the five boroughs for each year in order of decreasing values (from
Manhattan to Staten Island). Originally a spaghetti plot was used, however, it felt
worthwhile to diversify the analytical methods and emphasize the differences 
between boroughs, not just the trends.

  Figure 1 illustrates the HIV diagnosis rate by borough over time. It can be
observed that, over time, the HIV diagnosis rates have decreased in all of the
boroughs, except for Staten Island. For this particular borough, the HIV diagnosis
rate actually increased in the years 2013 and 2016. Additionally, there is also 
an observed spike in diagnosis rates in the year 2014 for Brooklyn. 

Also observed, over time, the Manhattan borough has the highest HIV diagnosis 
rate of all the NYC boroughs while Staten Island has the least. However, in 2015,
the Bronx has a higher number of HIV diagnosis rates compared to Manhattan. Lastly,
the Queens borough appears to have a relatively static change in HIV diagnosis
rates.


**Subgroup Visualizations**
```{r}
hiv = hiv_data %>%
  filter(borough == "All") %>%
  select(year, race, hiv_diagnosis_rate, gender) %>%
  filter(!race == "All")

hiv %>%
  ggplot(aes(x = year, y = hiv_diagnosis_rate, color = race, group = race)) +
  geom_line() +
  facet_grid(~gender) +
   labs(
    title = "HIV Diagnosis Rates Over Time by Race",
    x = "Year",
    y = "HIV Diagnosis Rates",
    caption = "Figure 3. Data from the NYCDOHMH HIV/AIDS Surveillance Annual Report"
  )
```

  In order to assess the HIV diagnosis rates by race, the 'All' observations within 
the borough variable was kept in the dataset (Figure 3). This was done to ensure 
that the output sought was not confounded by results specific to each borough, 
but rather to all boroughs. The variables year, race, gender and HIV diagnosis 
rates were spepcifically selected for this analysis. Additonally, being that the 
race variable was being visually assessed, the 'All' observations within the 
variable was filtered out. We thought it would have been interesting to observe 
the differences in HIV diagnosis rates by race and gender and, as such, the 
spaghetti plot was faceted by gender.

  Figure 3 illustrates the HIV diagnosis rates by race and faceted by gender over
time. It can be observed that, regardless of gender, Blacks have the highest 
numbers of diagnosis rates with Latinos/Hispanics having the second highest rates
over the 6-year period. Furthermore, Asian/Pacific Islander appears to have the 
lowest numbers of HIV diagnosis rates. There is also an interesting peak observed
among the race Other/Unknown observed in the year 2013, that cannot be explained.
Additionally, there is also an observed decrease in the rates of HIV diagnosis 
rates in the Black and Other/Unknown raceson when not looking at differentiation 
by gender.
  
  When visually assessing the plot by gender, males have a much higher HIV 
diagnosis rates compared to females. Looking more closely, the diagnosis rates of
Black males are significantly larger compared to Black females with the same 
distinction being made in all races, though not as wide. In Black males and 
females, there is a decrease in IV diagnosis rates over time. There is also a 
slight decrease in the Latino/Hispanic population in females, but an increase in 
the males. Furthermore, while there is a decrease in HIV diagnosis rates within 
White and Asian/Pacific Islander males, in females, there appears to be no 
linear change over the 6-year period.

```{r}
hiv_age =
  hiv_data %>%
  filter(age != "All", borough == "All", uhf == "All")

#HIV diag rate over time by age groups

hiv_age %>% 
  ggplot(aes(year, hiv_diagnosis_rate)) +
  geom_line(aes(group = age, color = age)) + 
  facet_wrap(~gender) +
  labs(
    title = "HIV Diagnosis Rates Over Time by Age Group",
    x = "Year",
    y = "HIV Diagnosis Rates",
    caption = "Figure 4. Data from the NYCDOHMH HIV/AIDS Surveillance Annual Report"
  )
```

  In Figure 4, the data was first filtered to obtain rows that only contained 
values for gender and age. Then, HIV diagnosis rates over time was plotted as a 
line graph, grouped by age groups and colored accordingly. To see if the rate
distribution among age groups differed by gender, the plot was faceted by gender 
as well.
  Similar to the observations made in Figure 3, males have a disproportionately
higher HIV diagnosis rate compared to females. In Figure 4, however, within 
females, there appears to be a relatively similar distribution of HIV diagnosis
rates between age groups over time. Conversely, among males, the diagnosis rates
differs drastically by age groupp. The 20-29 age group has the highest HIV 
diagnosis rates over time.Individuals within the 60+ age group has the lowest HIV
diagnosis rates over time. On a whole, for the age groups 20-29, 30-39, and 40-49,
there appears to be a decrease in HIV diagnosis rates. Tor the remaining age 
groups, the trend appears to remain constant at the year 2015.




**Additional Statistical Analysis**

Our exploratory data visualizations indicated that there are strong disparities in HIV Infection Rates amongst the five boroughs, various age groups and genders. As a result, we decided to fit a multiple linear regression model to compute parameter estimates and to see whether these disparities are statistically significant at the 5% level of significance.

```{r}
hiv_clean = hiv_data %>% 
  janitor::clean_names() %>% 
  filter(borough != "All")%>%
  filter(gender != "All")%>%
  filter(age != "All") %>% #Filtering out aggregate variables
  mutate(borough = forcats::fct_relevel(borough, c("Manhattan", "Bronx", "Brooklyn", "Queens", "Staten Island"))) 
#Reordering to make Manhattan the reference borough.

linear_fit = lm(hiv_diagnosis_rate ~ gender + borough + age, data = hiv_clean) 

linear_clean = linear_fit %>%
  broom::tidy() %>% 
  mutate(term = str_replace(term, "^borough", "Borough: ")) %>% 
  mutate(term = str_replace(term, "^gender", "Gender: ")) %>% 
  mutate(term = str_replace(term, "^age", "Age Group: "))

linear_clean %>% 
  knitr::kable(digits = 3)
```

Notable results from our multiple linear regression analysis include:

* Compared to Manhattan, all boroughs have a statistically significantly different HIV infection rate (p<0.01), adjusting for gender and age group. In particular, on average, the Bronx has a 0.330 percentage point higher HIV infection rate compared to Manhattan, adjusting for gender and age group. 
* On average, males have a 38.9 percentage point higher HIV infection rate compared to females, adjusting for borough and age group.
* On average, compared to 13-19 year olds, 20-29 year olds have a 42.6 percentage point higher HIV infection rate (p<0.01), adjusting for gender and borough. With age, the mean difference in HIV infection rate appears to decrease, adjusting for gender and borough. 

**Discussion**

**Data Can Be Found Here**
