---
title: "Final_Report"

author: "Noah Kreski(Ntk2109), Briana Lettsome(Bjl2150), Sunny Siddique (as5457), 
Bryce Wong (baw2160), David Yu (ky2288)"
date: "December 6th, 2018"
output: github_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(ggridges)

hiv_data = read_csv(file = "./data/DOHMH_HIV_AIDS_Annual_Report.csv") %>%
  janitor::clean_names()
```

##**Introduction**




## **Methods**

Data: Source, scraping method, cleaning, etc.

  The data that was utilized for this project was the New York City Department of
Health and Mntal Hygeine HIV/AIDS Surveillance Annual Report. The dataset was 
downloaded from NYCOpen Data, an open website that provides information and data to 
New Yorkers by the NYC government, an imported into R as a csv file. 
  Within this particular datThe data contained within the dataset were collected between the years 2011 through 2016. The HIV/AIDS surveillance Annual Report contains eighteen variables, including
year, age, race, HIV diagnoses, deaths, and percent viral suppression. The variables used within the visualizations are a followed:





## **Exploratory analysis**

: Visualizations, summaries, and exploratory statistical analyses. Justify the steps you took, and show any major changes to your ideas.

###*Big Picture Visualizations*

###*Subgroup Visualizations*
```{r}
hiv = hiv_data %>%
  filter(borough == "All") %>%
  select(year, race, hiv_diagnosis_rate, gender) %>%
  filter(!race == "All")

hiv %>%
  ggplot(aes(x = year, y = hiv_diagnosis_rate, color = race, group = race)) +
  geom_line() +
  facet_grid(~gender) +
   labs(
    title = "HIV Diagnosis Rates by Race",
    x = "Year",
    y = "HIV Diagnosis Rates",
    caption = "Data from the NYCDOHMH"
  )
```

  In order to assess the HIV diagnosis rates by race, the 'All' observations within 
the borough variable was kept in the dataset (Figure 3). This was done to ensure 
that the output we sought was not confounded by results specific to each borough, 
but rather to all boroughs. The variables year, race, gender and HIV diagnosis 
rates were spepcifically selected for this analysis. Additonally, being that the 
race variable was being visually assessed, the 'All' observations within the 
variable was filtered out. We thought it would have been interesting to observe 
the differences in in HIV diagnosis rates by race and gender and, a such, the 
spaghetti plot was faceted by gender.

  Figure 3 illustrates the HIV diagnosis rates by race and faceted by gender over
time. It can be observed that, regardless of gender, Blacks have the highest 
numbers of diagnosis rates with Latinos/Hispanics having the second highest rates
over the 6-year period. Furthermore, Asian/Pacific Islander appears to have the 
lowest numbers of HIV diagnosis rates. There is also an interesting peak observed
among the race Other/Unknown observed in the year 2013, that cannot be explained.
Additionally, there is also an observed decrease in the rates of HIV diagnosis 
rates in the Black and Other/Unknown raceson when not looking at differentiation 
by gender.
  
  When visually assessing the plot by gender, males have a much higher HIV 
diagnosis rates compared to females. Looking more closely, the diagnosis rates of
Black males are significantly larger compared to Black females with the same 
distinction being made in all races, though not as wide. In Black males and 
females, there is a decrease in IV diagnosis rates over time. There is also a 
slight decrease in the Latino/Hispanic population in females, but an increase in 
the males. Furthermore, while there is a decrease in HIV diagnosis rates within 
White and Asian/Pacific Islander males, in females, there appears to be no 
linear change over the 6-year period.

##**Results**




##**Discussion**


