---
title: "Data Science I: Final Report"
date: "December 6th, 2018"
output: github_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(ggridges)

hiv_data = read_csv(file = "./data/DOHMH_HIV_AIDS_Annual_Report.csv") %>%
  janitor::clean_names()
```

**Project Title**

Identification And Analysis of Trends And Disparities In HIV/AIDS For NYC

**Group Members**

* Bryce Wong (baw2160)
* Sunny Siddique (as5457)
* David Yu (ky2288)
* Noah Kreski (Ntk2109)
* Briana Lettsome (Bjl2150)


**Motivation / Related Work**

According to the the past decade or so, improvements in drug therapies and prevention intervations have led to a decreased number of AIDS cases. In June, 2014, New York Governor Andrew Cuomo announced a new strategy designed to "End the Epidemic" in the state through a combination of increased HIV screening and testing, promotion of Pre-exposure Prophylaxis (PrEP) and improved linkage and retention in clinical care for HIV-positive persons. Since then, the New York City Department of Health and Mental Hygiene (DOHMH) reported a decline in new infections of nearly 8 percent annually or approximately 15% since the campaign was launched.

Despite efforts to reduce the number of new infections, the HIV Surveillance Annual Report presented a number of key points that motivated our decision to analyze HIV infections data. Amongst them, the most noteworthy are:

* Poor New Yorkers, as well as African-American and Hispanic New Yorkers, bear a disproportionate burden of illness and premature death.
* Poor health is concentrated in certain New York City neighborhoods.
* Factors associated with poor health, such as poor access to medical care, unhealthy behaviors, and poor living conditions, are more common among certain economic and racial/ethnic groups.

**Initial Questions**

* Are there disparities in HIV infections / diagnoses based on NYC neighborhoods, if so, which neighborhoods have the highest incidence of HIV infections? Which neighborhoods have the lowest incidence?
* Are there disparities in common demographic factors, including age, gender and race? 
* Does HIV related death rate appear elevated for certain social groups? What about non-HIV related deaths?
* Are there specific observable trends in "percent viral suppression" over the past few years? If so, do the trends appear comparable to the DOHMH interventions following Governor Cuomo's announcements in 2014?



**Methods**

Data: Source, scraping method, cleaning, etc.

  The data that was utilized for this project was the New York City Department of
Health and Mental Hygeine (NYCDHMH) HIV/AIDS Surveillance Annual Report. This
dataset is contains HIV/AIDS-related data on individuals thirteen years or older 
in all NYC boroughs as well as throughou the city (DOHMH HIV/AIDS Annual Report).
The dataset was downloaded from NYCOpen Data, an open website that provides
information and data to New Yorkers by the NYC government. Once downloaded, the 
raw data was imported into Rstudio as a csv file. 
  Within this particular data there are The data contained within the dataset were
collected between the years 2011 through 2016. The HIV/AIDS surveillance Annual
Report contains eighteen variables, including year, age, race, HIV diagnoses,
deaths, and percent viral suppression and percent linked to care. The variable
outcomes used within the visualizations are defined as followed:

* Death rate types = 

* HIV diagnosis rate = "HIV diagnoses per 100,000 NYC population using annual
intercensal population estimates"

* Percent viral suppression = "Proportion of people living with diagnosed HIV
infection 13 years of age or older with at least one viral load test during the
calendar year whose last HIV viral load value was ≤200 copies/mL."


###**Exploratory analysis**


**Big Picture Visualizations**

```{r}
hiv_year_data = hiv_data %>% 
  filter(
    borough == "All",
    uhf == "All",
    gender == "All",
    age == "All",
    race == "All"
  )

all_death_rates = hiv_year_data %>% 
  gather(key = death_rate_type, value = death_rate_nums, death_rate:non_hiv_related_death_rate) %>% 
  filter(death_rate_nums != 99999)

stacked_death_plot = all_death_rates %>% 
  filter(death_rate_type != "death_rate")

ggplot(stacked_death_plot, aes(x = year, y = death_rate_nums, group = death_rate_type)) +   
  geom_col(aes(fill = death_rate_type)) +
  labs(
    title = "Death Rate Types Over the Years",
    x = "Year",
    y = "Rate",
    Caption = "Figure 1. Data from the NYCDOHMH HIV/AIDS Surveillance Annual Report"  )
```
  
  Data was first filtered so that the resulting data frame included only
observations that aggregated results across all boroughs, UHFs, genders, ages, 
and races. This data frame was then tidied so that death rate type (i.e.
"HIV-related death rate," "Non-HIV-related death rate," and overall "death rate")
was a single variable with the accompanying rate value in a separate variable
column. Any observation that had a rate value of "99999" was filtered out, as this
represented an observation where rate value was not calculated. Any observation
where the death rate type was the total death rate was also filtered out. A 
stacked bar graph was then constructed with rate values plotted on the y-axis and
year on the x-axis. Each bar represents the overall death rate per a year, 
segmented by death rate type ("HIV-related" and "Non-HIV-related").  
  
  Figure 1 illustrates the death rate types over the years. According to the 
stacked bar graph, there is a decreasing trend in HIV-related death types over the
years. Furthermore, there are more non-HIV-related deaths over the years.
Interestingly, however, these deaths are also decreasing over time. 

```{r}
# Developing the code to visually assess HIV diagnosis rates over time by borough

HIV_diagnosis_data = hiv_data%>%
               janitor::clean_names()%>%
               filter(uhf %in% c("All"))%>%
               filter(borough != "All")%>%
               filter(gender == "All")%>%
               filter(age == "All")

# Creating a plot to visually assess the above dataframe.

HIV_diagnosis_data%>%
            ggplot(aes(x = year, y = hiv_diagnosis_rate, group = borough, 
                       color = borough))+geom_point() + geom_line() +
  labs(
    title = "HIV Diagnosis Rate by Borough Over time",
    x = "Year",
    y = "HIV Diagnosis Rates",
    caption = "Figure 2. Data from the NYCDOHMH HIV/AIDS Surveillance Annual Report"
  )
```

   In Figure 2, The original dataset was filtered down to get the summary data by
borough for each year. Then, a bar plot was used for the HIV diagnosis rates,
clustered by the five boroughs for each year in order of decreasing values (from
Manhattan to Staten Island). Originally a spaghetti plot was used, however, it felt
worthwhile to diversify the analytical methods and emphasize the differences 
between boroughs, not just the trends.

  Figure 2 illustrates the HIV diagnosis rate by borough over time. It can be
observed that, over time, the HIV diagnosis rates have decreased in all of the
boroughs, except for Staten Island. For this particular borough, the HIV diagnosis
rate actually increased in the years 2013 and 2016. Additionally, there is also 
an observed spike in diagnosis rates in the year 2014 for Brooklyn. 

  Also observed, over time, the Manhattan borough has the highest HIV diagnosis 
rate of all the NYC boroughs while Staten Island has the least. However, in 2015,
the Bronx has a higher number of HIV diagnosis rates compared to Manhattan. Lastly,
the Queens borough appears to have a relatively static change in HIV diagnosis
rates.



**Subgroup Visualizations**
```{r}
hiv = hiv_data %>%
  filter(borough == "All") %>%
  select(year, race, hiv_diagnosis_rate, gender) %>%
  filter(!race == "All")

hiv %>%
  ggplot(aes(x = year, y = hiv_diagnosis_rate, color = race, group = race)) +
  geom_line() +
  facet_grid(~gender) +
   labs(
    title = "HIV Diagnosis Rates Over Time by Race",
    x = "Year",
    y = "HIV Diagnosis Rates",
    caption = "Figure 3. Data from the NYCDOHMH HIV/AIDS Surveillance Annual Report"
  )
```

  In order to assess the HIV diagnosis rates by race, the 'All' observations within 
the borough variable was kept in the dataset (Figure 3). This was done to ensure 
that the output sought was not confounded by results specific to each borough, 
but rather to all boroughs. The variables year, race, gender and HIV diagnosis 
rates were spepcifically selected for this analysis. Additonally, being that the 
race variable was being visually assessed, the 'All' observations within the 
variable was filtered out. We thought it would have been interesting to observe 
the differences in HIV diagnosis rates by race and gender and, as such, the 
spaghetti plot was faceted by gender.

  Figure 3 illustrates the HIV diagnosis rates by race and faceted by gender over
time. It can be observed that, regardless of gender, Blacks have the highest 
numbers of diagnosis rates with Latinos/Hispanics having the second highest rates
over the 6-year period. Furthermore, Asian/Pacific Islander appears to have the 
lowest numbers of HIV diagnosis rates. There is also an interesting peak observed
among the race Other/Unknown observed in the year 2013, that cannot be explained.
Additionally, there is also an observed decrease in the rates of HIV diagnosis 
rates in the Black and Other/Unknown raceson when not looking at differentiation 
by gender.
  
  When visually assessing the plot by gender, males have a much higher HIV 
diagnosis rates compared to females. Looking more closely, the diagnosis rates of
Black males are significantly larger compared to Black females with the same 
distinction being made in all races, though not as wide. In Black males and 
females, there is a decrease in IV diagnosis rates over time. There is also a 
slight decrease in the Latino/Hispanic population in females, but an increase in 
the males. Furthermore, while there is a decrease in HIV diagnosis rates within 
males of the White and Asian/Pacific Islander racial groups, in females, there
appears to be no linear change over time.

```{r}
split_by_race = hiv_data %>% 
  filter(
    age == "All",
    borough == "All",
    uhf == "All",
    race != "All"
    ) 

ggplot(split_by_race, aes(x = year, y = percent_viral_suppression, group = race)) +   
  geom_line(aes(color = race)) +
  facet_grid(~gender) +
  labs(
    title = "Percent Viral Suppression By Race Over the Years",
    x = "Year",
    y = "Percent Viral Suppression",
    caption = "Figure 4. Data from the NYCDOHMH HIV/AIDS Surveillance Annual Report"
  )
```

  Percent viral suppression by race and faceted by gender: Data was first filtered
so that the resulting data frame included only observations that aggregated
results across all boroughs, UHFs, and ages. This data frame also filtered out 
any observations that aggregated data across race, as we wanted to be able to
explore HIV outcomes by different races. There were no observations that
aggregated outcomes across all genders by each race, so I made the decision to 
leave these outcomes split by gender instead of manually combining findings for 
each gender. This meant I had to facet by gender in the resulting plot that explores
changes in an outcome for each race across years. A spaghetti plot was thus
constructed, faceted by gender, showing change in "percent viral suppression" on 
the y-axis for each race over years (on the x-axis).

   Figure 4 illiustrates the percent viral suppression by race over the years. 
In both plots, there is an overall increase in the percent viral supporession 
for each race over time. These results are strongly correlated with Figure 1, which 
shows HIV-related deaths have decreased over time. Over all, it appears that Whites
have the largest percent viral suppression compared to all races. 
  
  When faceted by gender, White males have the highest percent viral suppression
over time. Contrastingly, Black males have the lowest percent viral suppression, 
although the total percentage has increased over time. Furthermore, Black females
have higher suppression compared to their male counterparts. These results
correspond to the results observed in Figure 3. When looking at females, Whites 
and Asian/Pacidic Islanders have relatively similar percent viral supporession over
time. Interestingly, in females, the Other/Unknown race has the lowest percent
viral suppression.


```{r}
hiv_age =
  hiv_data %>%
  filter(age != "All", borough == "All", uhf == "All")

#HIV diag rate over time by age groups

hiv_age %>% 
  ggplot(aes(year, hiv_diagnosis_rate)) +
  geom_line(aes(group = age, color = age)) + 
  facet_wrap(~gender) +
  labs(
    title = "HIV Diagnosis Rates Over Time by Age Group",
    x = "Year",
    y = "HIV Diagnosis Rates",
    caption = "Figure 5. Data from the NYCDOHMH HIV/AIDS Surveillance Annual Report"
  )
```

  In Figure 5, the data was first filtered to obtain rows that only contained 
values for gender and age. Then, HIV diagnosis rates over time was plotted as a 
line graph, grouped by age groups and colored accordingly. To see if the rate
distribution among age groups differed by gender, the plot was faceted by gender 
as well.

  Similar to the observations made in Figure 3, males have disproportionately
higher HIV diagnosis rates compared to females. In Figure 4, however, within 
females, there appears to be a relatively similar distribution of HIV diagnosis
rates among age groups over time. Conversely, among males, the diagnosis rates
differs drastically by age groupp. The 20-29 age group has the highest HIV 
diagnosis rates over time.Individuals within the 60+ age group has the lowest HIV
diagnosis rates over time. On a whole, for the age groups 20-29, 30-39, and 40-49,
there appears to be a decrease in HIV diagnosis rates. For the remaining age 
groups, the trend appears to remain constant at the year 2015.


**Additional Statistical Analysis**

Our exploratory data visualizations indicated that there are strong disparities in HIV Infection Rates amongst the five boroughs, various age groups and genders. As a result, we decided to fit a multiple linear regression model to compute parameter estimates and to see whether these disparities are statistically significant at the 5% level of significance.

```{r}
hiv_clean = hiv_data %>% 
  janitor::clean_names() %>% 
  filter(borough != "All")%>%
  filter(gender != "All")%>%
  filter(age != "All") %>% #Filtering out aggregate variables
  mutate(borough = forcats::fct_relevel(borough, c("Manhattan", "Bronx", "Brooklyn", "Queens", "Staten Island"))) 
#Reordering to make Manhattan the reference borough.

linear_fit = lm(hiv_diagnosis_rate ~ gender + borough + age, data = hiv_clean) 

linear_clean = linear_fit %>%
  broom::tidy() %>% 
  mutate(term = str_replace(term, "^borough", "Borough: ")) %>% 
  mutate(term = str_replace(term, "^gender", "Gender: ")) %>% 
  mutate(term = str_replace(term, "^age", "Age Group: "))

linear_clean %>% 
  knitr::kable(digits = 3)
```

Notable results from our multiple linear regression analysis include:

* Compared to Manhattan, all boroughs have a statistically significantly different HIV infection rate (p<0.01), adjusting for gender and age group. In particular, on average, the Bronx has a 0.330 percentage point higher HIV infection rate compared to Manhattan, adjusting for gender and age group. 
* On average, males have a 38.9 percentage point higher HIV infection rate compared to females, adjusting for borough and age group.
* On average, compared to 13-19 year olds, 20-29 year olds have a 42.6 percentage point higher HIV infection rate (p<0.01), adjusting for gender and borough. With age, the mean difference in HIV infection rate appears to decrease, adjusting for gender and borough. 

**Discussion**

**Data Can Be Found Here**
