---
title: "Final_Report"

author: "Noah Kreski(Ntk2109), Briana Lettsome(Bjl2150), Sunny Siddique (as5457), 
Bryce Wong (baw2160), David Yu (ky2288)"
date: "December 6th, 2018"
output: github_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(ggridges)

hiv_data = read_csv(file = "./data/DOHMH_HIV_AIDS_Annual_Report.csv") %>%
  janitor::clean_names()
```

##**Introduction**




## **Methods**

Data: Source, scraping method, cleaning, etc.

  The data that was utilized for this project was the New York City Department of
Health and Mntal Hygeine HIV/AIDS Surveillance Annual Report. The dataset was 
downloaded from NYCOpen Data, an open website that provides information and data to 
New Yorkers by the NYC government, an imported into R as a csv file. 
  Within this particular datThe data contained within the dataset were collected between the years 2011 through 2016. The HIV/AIDS surveillance Annual Report contains eighteen variables, including
year, age, race, HIV diagnoses, deaths, and percent viral suppression. The variables used within the visualizations are a followed:





## **Exploratory analysis**

: Visualizations, summaries, and exploratory statistical analyses. Justify the steps you took, and show any major changes to your ideas.

###*Big Picture Visualizations*

###*Subgroup Visualizations*
```{r}
hiv = hiv_data %>%
  filter(borough == "All") %>%
  select(year, race, hiv_diagnosis_rate, gender) %>%
  filter(!race == "All")

hiv %>%
  ggplot(aes(x = year, y = hiv_diagnosis_rate, color = race, group = race)) +
  geom_line() +
  facet_grid(~gender) +
   labs(
    title = "HIV Diagnosis Rates by Race",
    x = "Year",
    y = "HIV Diagnosis Rates",
    caption = "Data from the NYCDOHMH"
  )
```

In order to assess the HIV diagnosis rates by race, the 'All' observations within 
the borough variable was kept in the dataset (Figure 3). This was done to ensure 
that the output we sought was not confounded by results specific to each borough, 
but rather to all boroughs. The variables year, race, gender and HIV diagnosis 
rates were spepcifically selected for this analysis. Additonally, being that the 
race variable was being visually assessed, the 'All' observations within the 
variable was filtered out.

##**Results**




##**Discussion**


